Reanalysis Step 7: Missingness Pattern Analysis

### 2025-09-01 reanalysis-7-missingness_patterns.qmd
fix ai idiocy

for (var in c("baseline_VA_logMAR", "CCI", "DCSI")) {
  if (var %in% available_vars) {
    # Get which observations were originally missing
    originally_missing <- is.na(original_data[[var]])

    # Calculate mean across imputations for originally missing
    if (sum(originally_missing) > 0) {
      means_missing <- sapply(imputed_datasets, function(df) {
        mean(df[[var]][originally_missing[df$cohort_id %in% included_ids]], na.rm = TRUE)
      })

      # Calculate mean for originally complete
      means_complete <- sapply(imputed_datasets, function(df) {
        mean(df[[var]][!originally_missing[df$cohort_id %in% included_ids]], na.rm = TRUE)
      })

      imputation_summary <- rbind(imputation_summary, data.frame(
        Variable = var,
        Originally_Complete = sprintf("%.3f (%.3f)", 
                                     mean(means_complete), sd(means_complete)),
        Imputed_for_Missing = sprintf("%.3f (%.3f)", 
                                      mean(means_missing), sd(means_missing)),
        Difference = sprintf("%.3f", mean(means_missing) - mean(means_complete))
      ))
    }
  }
}

>>>>

# Get IDs of included patients (those not excluded)
id_var <- if("cohort_id" %in% names(original_data)) "cohort_id" else "e_mrn_deidentified"
included_data <- original_data[!original_data$excluded, ]
included_ids <- unique(included_data[[id_var]])

for (var in c("baseline_VA_logMAR", "CCI", "DCSI")) {
  if (var %in% names(original_data)) {  # Check if variable exists in original data
    # Get which observations were originally missing
    originally_missing <- is.na(original_data[[var]])

    # Calculate mean across imputations for originally missing
    if (sum(originally_missing, na.rm = TRUE) > 0) {
      means_missing <- sapply(imputed_datasets, function(df) {
        # Match IDs to find corresponding rows in imputed data
        imputed_ids <- df[[id_var]]
        matching_rows <- imputed_ids %in% included_ids
        missing_and_included <- originally_missing[match(imputed_ids, original_data[[id_var]])] & matching_rows

        if (sum(missing_and_included, na.rm = TRUE) > 0) {
          mean(df[[var]][missing_and_included], na.rm = TRUE)
        } else {
          NA
        }
      })

      # Calculate mean for originally complete
      means_complete <- sapply(imputed_datasets, function(df) {
        imputed_ids <- df[[id_var]]
        matching_rows <- imputed_ids %in% included_ids
        complete_and_included <- !originally_missing[match(imputed_ids, original_data[[id_var]])] & matching_rows

        if (sum(complete_and_included, na.rm = TRUE) > 0) {
          mean(df[[var]][complete_and_included], na.rm = TRUE)
        } else {
          NA
        }
      })

      # Remove NAs before calculating statistics
      means_missing <- means_missing[!is.na(means_missing)]
      means_complete <- means_complete[!is.na(means_complete)]

      if (length(means_missing) > 0 && length(means_complete) > 0) {
        imputation_summary <- rbind(imputation_summary, data.frame(
          Variable = var,
          Originally_Complete = sprintf("%.3f (%.3f)", 
                                       mean(means_complete), sd(means_complete)),
          Imputed_for_Missing = sprintf("%.3f (%.3f)", 
                                        mean(means_missing), sd(means_missing)),
          Difference = sprintf("%.3f", mean(means_missing) - mean(means_complete))
        ))
      }
    }
  }
}
